---
title: "kickstarter"
---

For the purpose of this exercise, three R packages will be used. They are tidyverse, jsonlite and SmartEDA.

::: callout-note
You are required to install the R packages above, if necessary, before continue to the next step
:::

In the code chunk below, p_load() of pacman package is used to load the R packages into R environment

```{r}
pacman:: p_load(tidyverse, jsonlite, tidygraph, ggraph)
```

## Importing Graph Data

For the purpose of this exercise, *mc.json* file will be used. Before getting started, you should have the data set in the *data* sub-folder

::: Callout-warning
This is a large data set. Please ensure that you have added a line `data/` un the *.gitignore* file.
:::

In the code chunk below, fromJSON() of jsonlite package is used to import mc2.json file into R and save the output object

```{r}
write_rds
mcl_data <- fromJSON("VAST_Challenge_2026_MC1/MC1_final_00.json")
```

```{r}
write_rds(mc1_data, "data/MC1/mc1_data.rds")
```

```{r}
mc_data <- fromJSON("VAST_Challenge_2026_MC2/MC2 data.json")
org_chart <- fromJSON("VAST_Challenge_2026_MC2/org_chart.json")

```

```{r}
mc2_df <- as.data.frame(mc2_data$events)
```

```{r}
write_rds(mc2_df, "data/mc2_df.rds")

```

```{r}
write_rds(FILAH, "data/MC2/FILAH.rds")
write_rds(journalist, "data/MC2/journalist.rds")
write_rds(road_map, "data/MC2/road_map.rds")
write_rds(TROUT, "data/MC2/TROUT.rds")
```

```{r}
MC3_graph <- fromJSON("MC3_release/MC3_graph.json")
MC3_schema <- fromJSON("MC3_release/MC3_schema.json")
```

The 
















