---
title: "Hands-on Exercise 1"
author: "Javier Phang"
date: "April 15, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
freeze: false
---

# Getting Started

## Install and launch R packages.

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

### 1.3.1 R Graphics VS ggplot

```{r}
hist(exam_data$MATHS)
```

### 1.5 + 1.6 Essential Grammatical Elements in ggplot2

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS))
```

## 1.7 Essential Grammatical Elements in ggplot2: geom
- Geometric objects are the actual marks we put on a plot. Examples include:

  - geom_point for drawing individual points (e.g., a     scatter plot)
  - geom_line for drawing lines (e.g., for a line charts)
  - geom_smooth for drawing smoothed lines (e.g., for simple trends or approximations)
  - geom_bar for drawing bars (e.g., for bar charts)
  - geom_histogram for drawing binned values (e.g. a histogram)
  - geom_polygon for drawing arbitrary shapes
  - geom_map for drawing polygons in the shape of a map! (You can access the data to use for these maps by using the map_data() function).


### 1.7.1 Geometric Objects: geom_bar

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```

### 1.7.2 Geometric Objects: geom_dotplot

-   In a dot plot, the width of a dot corresponds to the bin width (or maximum width, depending on the binning algorithm), and dots are stacked, with each dot representing one observation.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

-   scale_y_continuous() is used to turn off the y-axis, and binwidth argument is used to change the binwidth to 2.5.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

### 1.7.3 Geometric Objects: geom_histogram()

-   In the code chunk below, geom_histogram() is used to create a simple histogram by using values in MATHS field of exam_data.

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
    geom_histogram()
```

### 1.7.4 Modifying a geometric object by changing geom()

-   **bins** argument is used to change the number of bins to 20
-   **fill** argument is used to shade the histogram with light blue color
-   **color** argument is used to change the outline colour of the bars to black

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  
```

### 1.7.5 Modifying a geometric object by changing aes()

-   The code chunk below changes the interior colour of the histogram (i.e. fill) by using sub-group of aesthetic().

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

### 1.7.6 Geometric Objects: geom-density()

-   geom-density() computes and plots kernel density estimate, which is a smoothed version of the histogram.

-   It is a useful alternative to the histogram for continuous data that comes from an underlying smooth distribution.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()           
```

-- plotting two kernel density lines

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

### 1.7.7 Geometric Objects:geom_boxplot

-   geom_boxplot() displays continuous value list. It visualises five summary statistics (the median, two hinges and two whiskers), and all “outlying” points individually.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            
```

-   Notches are used in box plots to help visually assess whether the medians of distributions differ. If the notches do not overlap, this is evidence that the medians are different.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

### 1.7.8 Geometric Objects: geom_violin

-   geom_violin is designed for creating violin plot. Violin plots are a way of comparing multiple data distributions. With ordinary density curves, it is difficult to compare more than just a few distributions because the lines visually interfere with each other. With a violin plot, it’s easier to compare several distributions since they’re placed side by side.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

### 1.7.9 Geometric Objects: geom_point()

-   geom_point() is useful for creating scatterplot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

### 1.7.10 *geom* objects can be combined

-   The code chunk below plots the data points on the boxplots by using both geom_boxplot() and geom_point().

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

## 1.8 Essential Grammatical Elements in ggplot2: stat

The [Statistics functions](https://ggplot2.tidyverse.org/reference/index.html) statistically transform data, usually as some form of summary. For example:

-   frequency of values of a variable (bar graph)
    -   a mean
    -   a confidence limit
-   There are two ways to use these functions:
    -   add a `stat_()` function and override the default geom, or
    -   add a `geom_()` function and override the default stat.

### 1.8.1 Working with stat()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

### 1.8.2 Working with stat - the *stat_summary()* method
- The code chunk below adds mean values by using stat_summary() function and overriding the default geom.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)  
```

### 1.8.3 Working with stat - the geom() method
- The code chunk below adding mean values by using geom_() function and overriding the default stat.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)          
```

### 1.8.4 Adding a best fit curve on a scatterplot? 
- The scatterplot below shows the relationship of Maths and English grades of pupils. The interpretability of this graph can be improved by adding a best fit curve.
- geom_smooth() is used to plot a best fit curve on the scatterplot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

- The default smoothing method can be overridden as shown below

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

## 1.9 Essential Grammatical Elements in ggplot2: Facets
- Facetting generates small multiples (sometimes also called trellis plot), each displaying a different subset of the data. They are an alternative to aesthetics for displaying additional discrete variables. ggplot2 supports two types of factes, namely: facet_grid() and facet_wrap.

### 1.9.1 Working with facet_wrap()
- facet_wrap wraps a 1d sequence of panels into 2d. This is generally a better use of screen space than facet_grid because most displays are roughly rectangular.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

### 1.9.2 facet_grid() function
- facet_grid() forms a matrix of panels defined by row and column facetting variables. It is most useful when you have two discrete variables, and all combinations of the variables exist in the data.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

## 1.10 Essential Grammatical Elements in ggplot2: Coordinates
- The Coordinates functions map the position of objects onto the plane of the plot. There are a number of different possible coordinate systems to use, they are:

-   [`coord_cartesian()`](https://ggplot2.tidyverse.org/reference/coord_cartesian.html): the default cartesian coordinate systems, where you specify x and y values (e.g. allows you to zoom in or out).
-   [`coord_flip()`](https://ggplot2.tidyverse.org/reference/coord_flip.html): a cartesian system with the x and y flipped.
-   [`coord_fixed()`](https://ggplot2.tidyverse.org/reference/coord_fixed.html): a cartesian system with a "fixed" aspect ratio (e.g. 1.78 for a "widescreen" plot).
-   [`coord_quickmap()`](https://ggplot2.tidyverse.org/reference/coord_map.html)

### 1.10.1 Working with Coordinate

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

- The code chunk below flips the horizontal bar chart into vertical bar chart by using coord_flip().

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

### 1.10.2 Changing the y- and x-axis range
- The scatterplot on the right is slightly misleading because the y-aixs and x-axis range are not equal.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

- Fixed both the y-axis and x-axis range from 0-100.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

## 1.11 Essential Grammatical Elements in ggplot2: themes
- Themes control elements of the graph not related to the data. For example: 
  - background colour
  - size of fonts
  - gridlines
  - colour of labels
  
- Built-in themes include: - theme_gray() (default) - theme_bw() - theme_classic()

### 1.11.1 Working with theme
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

- A horizontal bar chart plotted using theme_classic().
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

- A horizontal bar chart plotted using theme_minimal().
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

