---
title: "Take Home Exercise 2: VAST Challenge 2026 Mini-Challenge 1"
author: "Javier Phang"
date: "25 May, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: false
---

# 1 **Overview**

In this exercise, we will be tackling [Mini-case 1](https://vast-challenge.github.io/2026/MC1.html) of [VAST Challenge 2026.](https://vast-challenge.github.io/2026/index.html)

## 1.1 **Background**

TenantThread is a property technology company that uses AI-powered systems to manage tenant engagement and communication processes. During a confidential merger project known as Project HarborCrest, strict embargo controls were implemented to prevent sensitive information from being released before the official announcement.

However, confidential merger-related information was leaked through AI-assisted communication channels prior to the embargo deadline. This raised concerns about whether the release was caused by deliberate actions, abnormal agent behaviour, or failures in the automated compliance system.

This mini-challenge focuses on analysing communication logs, agent interactions, and environmental events to reconstruct the sequence of events leading to the embargo breach.

## 1.2 **Objectives**

This analysis aims to:

1.  Reconstruct the sequence of events leading to the inappropriate release of confidential merger information by identifying key actions, causal relationships, communication patterns, and decision points involved in the embargo breach.
2.  Examine the behaviours and interactions of the systems, agents, and individuals involved to understand how the release bypassed the embargo enforcement mechanisms.
3.  Compare the behaviours leading to the release against prior historical behaviours to identify abnormal or unusual patterns.
4.  Detect potential leading indicators and prior instances where agents or systems behaved differently from their expected behaviours.
5.  Identify whether similar behavioural patterns had previously occurred and investigate why earlier incidents did not result in noticeable intervention or corrective action.
6.  Evaluate whether the inappropriate release was primarily caused by deliberate actions, abnormal agent behaviour, or weaknesses within the automated compliance and communication systems.

# 2 **Data Preparation and Initial Exploration**

This section examines the Mini-Challenge 1 dataset and outlines the data preparation and exploratory steps performed prior to the visual analytics process.

The raw JSON files were imported and transformed into structured tabular datasets to support downstream temporal, behavioural and network-based analysis. A preliminary exploratory analysis was carried out to gain a clearer understanding of the dataset's structure and relationships before moving on to a more in-depth investigation.

## 2.1 **Loading Packages**

The analysis primarily uses packages from the *tidyverse ecosystem* for data preprocessing, transformation, and exploratory analysis. Additional visual analytics libraries are incorporated to support network visualisation, temporal analysis, and interactive graphical exploration.

::: callout-note
You are required to install the R packages above, if necessary, before proceeding to the next step.
:::

In the code chunk below, *p_load()* of *pacman* package is used to load the R packages into R environment

```{r}
pacman::p_load(
  tidyverse, jsonlite, tidygraph, ggraph,
  knitr, lubridate, patchwork,
  igraph, ggiraph, readr
)
```

### 2.1.1 **Visual Analytics and Functions**

In addition to regular data wrangling packages, various specialized visual analytics libraries are included to facilitate network analysis, temporal investigation, and interactive visualization of the Mini-Challenge 1 dataset.

The `tidygraph`, `ggraph`, and `igraph` packages are mainly utilized for analyzing communication networks and visualizing relationships among agents and systems related to the embargo breach.

The lubridate library facilitates temporal analysis by handling timestamps, while `ggiraph` boosts interactivity in graphical results. The `patchwork` package serves to merge various visualizations into cohesive analytical layouts. Thus, these libraries enable exploratory analysis and visual storytelling to reveal behavioral trends and communication dynamics in the AI-supported business setting.

```{r}
packages <- tribble(
  ~Library, ~Description,
  "tidyverse", "Data preprocessing and wrangling.",
  "jsonlite", "Importing and parsing JSON datasets.",
  "tidygraph", "Graph and network analysis.",
  "ggraph", "Network visualisation.",
  "igraph", "Graph analytics and relationship analysis.",
  "lubridate", "Date-time and temporal analysis.",
  "patchwork", "Combining multiple visualisations.",
  "ggiraph", "Interactive graphical visualisations.",
  "knitr", "Dynamic report generation in Quarto."
)

knitr::kable(packages)
```

## 2.2 **Importing and Preparing MC1 Data**

For this exercise, the `MC1_final_00.json` dataset from Mini-Challenge 1 is used. Before getting started, the dataset should be stored in the `data/MC1/` sub-folder.

::: Callout-warning
This is a large data set. Please ensure that `data` has been added into the `.gitignore` file.
:::

In the code chunk below, `fromJSON()` from the `jsonlite` package is used to import the MC1 dataset into the R environment.

```{r}
mc1_data <- jsonlite::read_json(
  "MC1_final_00.json",
  simplifyVector = FALSE
)
```

The imported JSON object is then saved as an `.rds` file to improve loading efficiency for subsequent analysis.

```{r}
dir.create("data/MC1", recursive = TRUE, showWarnings = FALSE)

readr::write_rds(mc1_data, "data/MC1/mc1_data.rds")
```

The nested communication records were subsequently extracted and transformed into a structured tabular dataset suitable for downstream exploratory and visual analytics.

```{r}
communications <- map_dfr(mc1_data$rounds, function(round) {
  
  map_dfr(round$communications, function(msg) {
    
    tibble(
      round_hour = round$hour,
      message_id = msg$message_id,
      agent_id = msg$agent_id,
      agent_role = msg$agent_role,
      channel = msg$channel,
      message_type = msg$message_type,
      recipients = paste(msg$recipients, collapse = ", "),
      content = msg$content,
      timestamp = msg$timestamp
    )
  })
})
```

## 2.3 **Dataset Overview**

Before conducting visual analytics, an initial exploration of the transformed communication dataset was performed to understand its structure, dimensions, and variable composition. This helps identify the key fields available for analysing communication behaviour, agent interactions, and temporal activity patterns.

### Glimpse of Communication Dataset

The `glimpse()` function is used to provide a compact overview of the transformed communication dataset.

```{r}
glimpse(communications)
```

### Dataset Dimensions

```{r}
dataset_summary <- tibble(
  Measure = c("Number of communication records", 
              "Number of variables",
              "Number of unique agents",
              "Number of communication channels",
              "Number of simulation rounds"),
  Value = c(
    nrow(communications),
    ncol(communications),
    n_distinct(communications$agent_id),
    n_distinct(communications$channel),
    n_distinct(communications$round_hour)
  )
)

knitr::kable(dataset_summary)
```

::: callout-note
The transformed communication dataset contains message-level records extracted from the MC1 JSON file. Each row represents one communication event, including the sender, channel, message type, timestamp, and message content.
:::

## 2.4 **Data Quality Assessment**

An initial assessment of data quality was carried out to identify missing values and verify the completeness of the transformed communication dataset prior to executing subsequent visual analytics.

```{r}
missing_summary <- communications %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Missing_Count"
  )

knitr::kable(missing_summary)
```

::: callout-note
The transformed communication dataset does not contain substantial missing values and is sufficiently complete for downstream temporal, behavioural, and network-based analysis.
:::

# 3 Data Wrangling and Communication Engineering

This section engineers additional variables required to reconstruct the sequence of events, examine communication relationships, and identify behavioural indicators that may have contributed to the embargo breach. The engineered features are designed to support temporal, behavioural, and network-based visual analytics.

## 3.1 Engineering Temporal Communication Features

To reconstruct how the embargo breach unfolded, additional time-based variables were engineered from the original communication timestamps. These variables support event sequencing, communication timeline analysis, and identification of activity patterns before the inappropriate release.

```{r}
communications_clean <- communications %>%
  mutate(
    timestamp = ymd_hms(timestamp),
    round_hour = ymd_hms(round_hour),

    communication_date = as_date(timestamp),
    communication_hour = hour(timestamp),
    communication_day = wday(timestamp, label = TRUE),

    communication_period = case_when(
      communication_hour >= 6 & communication_hour < 12 ~ "Morning",
      communication_hour >= 12 & communication_hour < 18 ~ "Afternoon",
      communication_hour >= 18 & communication_hour < 24 ~ "Evening",
      TRUE ~ "Late Night"
    ),

    days_before_embargo_breach =
      as.numeric(as_date("2046-06-05") - communication_date),

    near_breach_window = case_when(
      days_before_embargo_breach <= 2 ~ "Critical Period",
      days_before_embargo_breach <= 5 ~ "Escalation Period",
      TRUE ~ "Normal Period"
    ),

    message_sequence = row_number()
  )
```

The engineered temporal variables are verified below.

```{r}
communications_clean %>%
  select(
    timestamp,
    communication_date,
    communication_hour,
    communication_day,
    communication_period,
    days_before_embargo_breach,
    near_breach_window,
    message_sequence
  ) %>%
  head()
```

::: callout-note
The `days_before_embargo_breach` variable ties the communication timeline to the breach date of June 5, 2046, facilitating subsequent analysis to contrast typical communication patterns with actions nearer to the embargo violation.
:::

## 3.2 Classifying Communication Scope and Exposure Risk

To investigate how embargoed information may have propagated across different communication environments, communication channels were grouped into broader communication scopes based on their risk of exposure.

The classification distinguishes between public-facing communication, private exchanges, internal coordination discussions, and side-channel interactions. This supports downstream analysis of whether sensitive information gradually transitioned from internal discussions into externally visible communication channels prior to the embargo breach.

The unique communication channels are first examined to understand the types of communication environments present in the dataset before defining broader communication scope categories.

```{r}
communications_clean %>%
  distinct(channel) %>%
  arrange(channel) %>%
  knitr::kable()
```

```{r}
communications_clean <- communications_clean %>%
  mutate(
    communication_scope = case_when(
      channel %in% c("official_post", "personal_post") ~ "Public",
      channel %in% c("one_on_one_chat") ~ "Private",
      channel %in% c("side_huddle") ~ "Side-channel",
      channel %in% c("comms_huddle") ~ "Internal Group",
      TRUE ~ "Other"
    ),

    exposure_risk_level = case_when(
      communication_scope == "Public" ~ "High Exposure",
      communication_scope == "Side-channel" ~ "Moderate Exposure",
      communication_scope == "Private" ~ "Low Exposure",
      communication_scope == "Internal Group" ~ "Controlled Exposure",
      TRUE ~ "Unknown"
    ),

    is_public_channel = communication_scope == "Public"
  )
```

The communication scope classifications are verified below.

```{r}
communications_clean %>% 
  count(
    channel,
    communication_scope,
    exposure_risk_level
  ) %>% 
  arrange(desc(n)) %>% 
  knitr::kable()
```

::: callout-note
The engineered communication scope categories support downstream investigation of how merger-related information may have propagated across communication environments with different levels of exposure risk prior to the embargo breach.
:::

## 3.3 Engineering Embargo and Merger-related Communication Indicators

To support downstream behavioural and anomaly analysis, further text-based indicators were developed to pinpoint communication records possibly linked to merger talks, embargo-related issues, and governance risks. 

Keyword-based indicators were created to identify mentions concerning Project HarborCrest, CivicLoom, enforcement of embargoes, discussions on compliance, and potential information leakage behaviour.

```{r}
communications_clean <- communications_clean %>%
  mutate(
    content_lower = str_to_lower(content),

    contains_project_terms = str_detect(
      content_lower,
      "harborcrest|civicloom|merger|rebrand|capital|deal"
    ),

    contains_embargo_terms = str_detect(
      content_lower,
      "embargo|judge|compliance|approved|blocked|deadline"
    ),

    contains_public_exposure_terms = str_detect(
      content_lower,
      "public|post|fle?x|release|leak|disclose|announcement"
    ),

    contains_pressure_terms = str_detect(
      content_lower,
      "pressure|urgent|concern|risk|legal|scrutiny|journal|algorithmiceviction"
    ),

    message_length = str_length(content),
    word_count = str_count(content, "\\S+"),

    risk_signal_score =
      as.integer(contains_project_terms) +
      as.integer(contains_embargo_terms) +
      as.integer(contains_public_exposure_terms) +
      as.integer(contains_pressure_terms),

    risk_signal_group = case_when(
      risk_signal_score >= 3 ~ "High Signal",
      risk_signal_score == 2 ~ "Moderate Signal",
      risk_signal_score == 1 ~ "Low Signal",
      TRUE ~ "No Signal"
)
  )
```

```{r}
communications_clean %>%
  count(risk_signal_group) %>%
  arrange(desc(n)) %>%
  knitr::kable()
```

::: callout-note
The `risk_signal_score` consolidates multiple case-specific indicators into a single behavioural signal. This supports downstream analysis of whether merger-related, embargo-related, public exposure, and pressure-related communication increased closer to the breach.
:::

## 3.4 **Preparing Communication Nodes and Edges**

To facilitate downstream network-oriented visual analytics, the transformed communication records were reshaped into node and edge formats appropriate for graph analysis.

The node table represents the communication participants within the simulation environment, whereas the edge table records the directional relationships between senders and receivers through various communication channels.

### Preparing Communication Edges

The communication recipients were subsequently separated to construct directional sender-recipient relationships suitable for network analysis.

```{r}
communication_edges <- communications_clean %>%
  separate_rows(recipients, sep = ", ") %>%
  mutate(
    source = str_trim(agent_id),
    target = str_trim(recipients)
  ) %>%
  filter(
    !is.na(source),
    !is.na(target),
    source != "",
    target != "",
    target != "ALL"
  ) %>%
  transmute(
    source,
    target,
    channel,
    communication_scope,
    exposure_risk_level,
    risk_signal_group,
    timestamp
  )
```

The structure of the communication edges is examined below.

```{r}
glimpse(communication_edges)
```

### Preparing Communication Nodes

The communication relationships were aggregated from both senders and recipients to ensure that all participants appearing in the edge table are represented in the network.

```{r}
communication_nodes <- tibble(
  node_id = unique(c(
    communication_edges$source,
    communication_edges$target
  ))
) %>%
  left_join(
    communications_clean %>%
      distinct(agent_id, agent_role) %>%
      rename(node_id = agent_id, role = agent_role),
    by = "node_id"
  ) %>%
  mutate(
    role = replace_na(role, "Recipient / System")
  )
```

The structure of the communication nodes is examined below..

```{r}
glimpse(communication_nodes)
```

### Preparing Weighted Communication Edges

The communication relationships were aggregated to calculate interaction intensity between sender-recipient pairs.

```{r}
communication_edges_weighted <- communication_edges %>%
  count(
    source,
    target,
    communication_scope,
    exposure_risk_level,
    risk_signal_group,
    name = "weight"
  ) %>%
  mutate(
    interaction_intensity = case_when(
      weight >= 10 ~ "High Interaction",
      weight >= 5 ~ "Moderate Interaction",
      TRUE ~ "Low Interaction"
    )
  )
```

```{r}
head(communication_edges_weighted)
```

The relationship between interaction intensity and behavioural risk signals is further examined below.

```{r}
communication_edges_weighted %>%
  count(
    interaction_intensity,
    risk_signal_group
  ) %>%
  arrange(desc(n)) %>%
  knitr::kable()
```

The overall size of the prepared network components is summarised below.

```{r}
tibble(
  Measure = c(
    "Number of communication nodes",
    "Number of communication edges",
    "Number of weighted communication edges"
  ),
  Value = c(
    nrow(communication_nodes),
    nrow(communication_edges),
    nrow(communication_edges_weighted)
  )
) %>%
  knitr::kable()
```

::: callout-note
The engineered node and edge structures support downstream investigation of communication flow, interaction intensity, and potential information propagation pathways associated with the embargo breach.
:::

## 3.5 **Constructing the Communication Network**

The prepared communication nodes and edges were subsequently combined into a directed communication network suitable for downstream graph visualization and network analysis.

The communication network records the directional flow of information among participants and enables further examination of communication trends, interaction strength, and possible pathways of information dissemination linked to the embargo violation.

```{r}
communication_network <- tbl_graph(
  nodes = communication_nodes,
  edges = communication_edges_weighted,
  directed = TRUE
)
```

The structure of the constructed communication network is examined below.

```{r}
communication_network
```

::: callout-note
The constructed communication network serves as the foundation for subsequent visual analytics involving communication flow, network centrality, behavioural escalation, and potential dissemination pathways related to the inappropriate release.
:::

# 4 **Exploratory Communication and Behaviour Analysis**

## 4.1 Are There Signs of Escalating Communication Before the Breach?

This section investigates whether communication activity exhibited unusual escalation patterns approaching the embargo breach on June 5, 2046.

The analysis focuses on identifying:
-   increases in communication activity,
-   focused interaction intervals,
-   increase in behavioral risk indicators,
-   and possible warning signs preceding the improper release.

These patterns might indicate early signs of communication instability, coordination stress, or declining embargo controls before the breach occurs.

::: panel-tabset

### Daily Communication Activity 

```{r}
communication_daily <- communications_clean %>%
  count(communication_date)

ggplot(
  communication_daily,
  aes(x = communication_date, y = n)
) +
  
  annotate(
    "rect",
    xmin = as.Date("2046-06-03"),
    xmax = as.Date("2046-06-05"),
    ymin = -Inf,
    ymax = Inf,
    alpha = 0.1,
    fill = "red"
  ) +
  
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  
  geom_vline(
    xintercept = as.Date("2046-06-05"),
    linetype = "dashed",
    colour = "red"
  ) +
  
  annotate(
    "text",
    x = as.Date("2046-06-05"),
    y = max(communication_daily$n, na.rm = TRUE),
    label = "Embargo Breach",
    hjust = 1.05,
    colour = "red"
  ) +
  
  scale_y_log10() +
  
  labs(
    title = "Daily Communication Activity Prior to the Embargo Breach",
    x = "Communication Date",
    y = "Number of Communications (Log Scale)"
  ) +
  
  theme_minimal()
```

The timeline provides an overview of how communication activity changed across the investigation period. Any increase in communication volume closer to the breach may indicate growing coordination pressure or operational escalation.

### Behavioural Risk Signals Over Time

```{r}
risk_signal_timeline <- communications_clean %>%
  count(
    communication_date,
    risk_signal_group
  )

ggplot(
  risk_signal_timeline,
  aes(
    x = communication_date,
    y = n,
    fill = risk_signal_group
  )
) +
  annotate(
    "rect",
    xmin = as.Date("2046-06-03"),
    xmax = as.Date("2046-06-05"),
    ymin = -Inf,
    ymax = Inf,
    alpha = 0.1,
    fill = "red"
  ) +
  geom_col(position = "stack") +
  geom_vline(
    xintercept = as.Date("2046-06-05"),
    linetype = "dashed",
    colour = "red"
  ) +
  labs(
    title = "Behavioural Risk Signals Leading Up to the Embargo Breach",
    subtitle = "Stacked daily risk signals highlight escalation before the inappropriate release",
    x = "Communication Date",
    y = "Number of Communications",
    fill = "Risk Signal"
  ) +
  theme_minimal()
```

::: callout-note
This section helps identify whether communication activity and behavioural risk signals intensified before the embargo breach. These patterns provide early evidence for assessing whether the inappropriate release was preceded by measurable warning signs.
:::

:::

## 4.2 How Did Communication Behaviour Shift Across Communication Environments?

This section investigates whether communication behaviour progressively shifted across different communication environments prior to the embargo breach on June 5, 2046.

The analysis focuses on examining changes in communication scope, exposure-risk escalation, and the potential breakdown of communication containment surrounding embargo-sensitive information.

### Communication Environment Distribution

```{r}
communication_scope_distribution <- communications_clean %>%
  count(
    near_breach_window,
    communication_scope
  ) %>%
  group_by(near_breach_window) %>%
  mutate(
    percentage = n / sum(n)
  )

ggplot(
  communication_scope_distribution,
  aes(
    x = near_breach_window,
    y = percentage,
    fill = communication_scope
  )
) +
  
  geom_col(
    position = "fill",
    width = 0.7
  ) +
  
  scale_y_continuous(
    labels = scales::percent
  ) +
  
  labs(
    title = "Communication Environment Distribution Across Breach Windows",
    subtitle = "Comparing communication environments before the embargo breach",
    x = "Breach Window",
    y = "Percentage of Communications",
    fill = "Communication Scope"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

The proportional communication scope distribution illustrates how communication environments evolved across different breach windows. Increasing concentrations of public-facing or side-channel communication closer to the embargo breach may indicate weakening containment surrounding sensitive merger-related discussions.

### Exposure-risk Communication Behaviour

```{r}
exposure_risk_heatmap <- communications_clean %>%
  count(
    near_breach_window,
    exposure_risk_level
  )

ggplot(
  exposure_risk_heatmap,
  aes(
    x = near_breach_window,
    y = exposure_risk_level,
    fill = n
  )
) +
  
  geom_tile(
    colour = "white",
    linewidth = 0.5
  ) +
  
  geom_text(
    aes(label = n),
    size = 4
  ) +
  
  scale_fill_gradient(
    low = "#dbe9f6",
    high = "#08306b"
  ) +
  
  labs(
    title = "Exposure-risk Communication Across Breach Windows",
    subtitle = "Comparing communication exposure levels before the embargo breach",
    x = "Breach Window",
    y = "Exposure-risk Level",
    fill = "Messages"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

The exposure-risk comparison highlights whether higher-risk communication environments became increasingly prevalent during the escalation and critical breach periods. Such behavioural changes may suggest weakening communication controls and increased exposure vulnerability prior to the embargo violation.

