---
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: slide
    smaller: true
    incremental: false

execute:
  eval: true
  echo: false
  warning: false
  message: false

freeze: false
---

```{r}
#| include: false

pacman::p_load(tidyverse,ggplot2,scales,patchwork,ggdist,ggridges,ggstatsplot)

```

```{r}
#| include: false

motor_raw <- read_csv(
  "Dataset of motor insurance portfolio.csv"
)
```

## Motor Insurance Portfolio Analytics {layout-valign="center" layout-align="center"}

### Executive Summary using the IDEA Framework

<br><br>

Javier Phang

ISSS608 Visual Analytics and Applications

Singapore Management University

<br>

16 May 2026

::: notes
This presentation summarises the key findings from the motor insurance portfolio analysis using the Interactive Data Exploration and Analysis framework. The analysis focuses on claim behaviour, underwriting profitability, portfolio risk exposure, and pricing adequacy across the motor insurance portfolio.
:::
